Endogenous Politics and the Design of Trade Institutions\thanks{I thank Kyle Bagwell, Mostafa Beshkar, Rick Bond, Chad Bown, Nuno Limao, Giovanni Maggi, Anson Soderbery, Bob Staiger, Maurizio Zanardi, Ben Zissimos and participants in seminars at SITE for very helpful comments and discussions.}

### 2018-03-03 draft.tex
recompile without weird title page for web posting
re-write abstract; replace intro with relevant slides from UNL talk

\documentclass[12pt,titlepage]{article}

@@

\begin{document}
\title{Endogenous Politics and the Design of Trade Institutions\thanks{I thank Kyle Bagwell, Mostafa Beshkar, Rick Bond, Renee Bowen, Chad Bown, Paola Conconi, David DeRemer, James Lake, Nuno Lim\~ao, Giovanni Maggi, Devashish Mitra, Peter Rosendorff, Bob Staiger, Maurizio Zanardi, Ben Zissimos and participants in seminars at McGill, Rochester, Ryerson, Wesleyan, ETSG, InsTED, the Midwest International Trade Meetings, PEIO and SITE for very helpful comments and discussions.}}
\author{Kristy Buzard\thanks{Syracuse University, Economics Department. Email: kbuzard@syr.edu. http://faculty.maxwell.syr.edu/kbuzard.}}
\date{\vskip-.1in \today}
\maketitle

\vskip.3in
\begin{center} {\bf Abstract} \end{center}

\begin{quote}
{\small Much of the literature on the design of trade policy institutions takes the political pressure that governments face as exogenous. This paper integrates exogenous and endogenous political pressure into a single model, showing that the presence of endogenous political pressure overturns important results and provides insight into observed features of trade agreements as well as rules of organizations like the WTO. Endogenous politics destroys a traditionally-defined escape clause's ability to provide flexibility in the face of political shocks when lobbies use the flexibility to seek rents. This can explain why WTO Safeguard use is conditioned on measurable economic indicators.}
\end{quote}

@@

\bigskip
\section{Introduction}
\input{intro}

>>>>

%\documentclass[12pt,titlepage]{article} %for submission to AEJ-micro
\documentclass[12pt]{article}

@@

\begin{document}
\title{Endogenous Politics and the Design of Trade Institutions\thanks{I thank Kyle Bagwell, Mostafa Beshkar, Rick Bond, Renee Bowen, Chad Bown, Paola Conconi, David DeRemer, James Lake, Nuno Lim\~ao, Giovanni Maggi, Devashish Mitra, Peter Rosendorff, Bob Staiger, Maurizio Zanardi, Ben Zissimos and participants in seminars at McGill, Rochester, Ryerson, Wesleyan, ETSG, InsTED, the Midwest International Trade Meetings, PEIO, SITE, and the University of Nebraska for very helpful comments and discussions.}}
\author{Kristy Buzard\thanks{Syracuse University, Economics Department. Email: kbuzard@syr.edu. http://faculty.maxwell.syr.edu/kbuzard.}}
\date{\vskip-.1in \today}

\maketitle

\vskip.3in
\begin{center} {\bf Abstract} \end{center}
\begin{quote}
{\small While most of the literature on the design of trade agreements and trading institutions takes the political pressure governments face to be exogenous, this paper endogenizes politics in a standard model for studying trade policy design questions. This permits a careful examination of the impact of assumptions about the policy-maker's objective function on conclusions about the role that trade agreements play. With this simple modeling framework, one can carefully distinguish between the dynamics induced by exogenous political shocks and endogenous incentives of political actors, unifying these two distinct strands of the literature. The modeling framework can also provide fuller answers to trade policy design questions by elucidating the interactions between exogenous and endogenous political forces. With applications to tariff caps and escape clauses, the modeling framework shows that important insights are missed if attention is restricted to exogenous political shocks. Most notably, endogenous politics destroys a traditionally-defined escape clause's ability to provide flexibility in the face of political shocks when lobbies use the flexibility to seek rents. This can explain why WTO Safeguard use is conditioned on measurable economic indicators.}
\end{quote}

@@

\bigskip
\section{Introduction}

\begin{enumerate}
\item Can trade agreements (TAs) be used to manipulate domestic lobbying incentives?
	\begin{itemize}
		\item Government objective function
	\end{itemize}
\vskip.1in
	\item What is the optimal design of various trade agreement properties?
	\begin{itemize}
		\item Exogenous vs. endogenous politics
	\end{itemize}
\end{enumerate}

Political Economy of Trade Institutions\\

With a few exceptions, TA design literature has taken political economy forces to be exogenous. I:
\begin{itemize}
	\item endogenize politics into a standard model for studying TA design questions
		\begin{itemize}
			\item use this to examine gov't objective
		\end{itemize}
	\item carefully distinguish between dynamics induced by exogenous and endogenous politics for
		\begin{itemize}
			\item base case with tariff caps
			\item tariff caps with escape clause
		\end{itemize}
	\item examine escape clause design when both exogenous and endogenous forces are present
\end{itemize}

Results
\begin{itemize}
	\item TAs may be used to manipulate domestic political actors (even with no long-run distortions)
	\item Simple modeling framework can capture results from models of both exogenous and endogenous politics
		\begin{itemize}
			\item For both tariff caps and escape clauses, endogenous politics changes outcomes dramatically
		\end{itemize}
	\item Standard, theoretical escape clause can't work in the presence of endogenous political pressure
		\begin{itemize}
			\item Points to real-world design of WTO Agreement on Safeguards
			\item May explain why escape clause has fallen out of use
		\end{itemize}
\end{itemize}

Objective Function

Restraining Political Pressure through TAs
\begin{itemize}
	\item Will TA be used to discourage lobbying? Depends on how gov't welfare varies in $\ga$
	\item With standard Baldwin-style objective function, welfare always increases with $\ga$
\[
  W = \mathit{CS}_X(\tau) + \ga \pi_X(\tau) + \mathit{CS}_Y(\tau^*) + \pi_Y(\tau^*) + \mathit{TR}(\tau)
\]

	\begin{itemize}
		\item Isomorphic to `Protection for Sale' objective function
	\end{itemize}

	\item If lobbying effort subtracted as cost from $W$, welfare no longer monotonic in $\ga$
		\begin{itemize}
			\item If weights must sum to 1, welfare also not monotonic in $\ga$
		\end{itemize}
\end{itemize}

Comparison to the Literature\\

Others derive non-monotonicity in lobbying effort/tariff

\begin{itemize}
	\item Maggi $\&$ Rodriguez-Clare (1998/2007): dynamic firm investment disortions
	\item Mitra (2002): lobbies pay investment cost to form
\end{itemize}

\vskip.15in

Here, I acknowledge the gov't objective function may be fundamentally non-monotonic

\begin{itemize}
	\item Achieve same results with simpler model
	\item Endogenous politics in a wider range of questions
	\item Can have both endogenous / exogenous at the same time $\Rightarrow$ unify the exogenous and endogenous politics literatures
\end{itemize}

%\input{intro}
